Manuscript_draft

### 2021-03-16 manuscript_draft.Rmd
include growth dominance coefficient information into discussion

---
title: "Manuscript_draft"
author: "Michael Jull and Hardy Griesbauer"
date: "10/12/2020"
output:
  word_document:
    reference_docx: wordTemplate2.docx
  html_document:
    df_print: paged
csl: cjfr.csl
bibliography: ep1162.bib
editor_options: 
  markdown: 
    wrap: 72
---

@@

To test the hypothesis that tree-level radial growth response over the
1994-2019 period varied with interactions between (i) tree diameter at
the time of final treatment (1994); (ii) treatment; and (iii) species,
we fit linear mixed effects models with log-transformed 1994-2019
diameter increment as the response variable, and treatment, species and
log-transformed 1992 diameter as interacting fixed effects. To account
for plot-level productivity differences, we included plot as a random
intercept effect. Models were fit using the 'lme4' package for R [\@].
We also fit a similar model with the 1994-2019 diameter increment
expressed as a percentage of 1994 diameter. Plots of residuals and fit
lines with orginal data were examined to evaluate model goodness of fit.
Variance explained for each model are reported using the approach
described in Nakagawa et al. [-@Nakagawa2013], as implemented in the
MuMIn package for R [@barton2009mumin]. Pairwise contrasts were
performed using the 'emmeans' package for R [\@] and Tukey HSD method of
*p* value adjustment was used to control for multiplicity. We report all
significant results at p\<0.05.

@@

# Filter only for trees that have measurements in 1992 and 2019 
treeList<- # assign variable
  X %>% 
  filter(TreeID!="New") %>% # remove new trees
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(Species %in% c("Spruce","Fir")) %>% # filter birch and fdi out of dataset
  filter(Status=="Live") %>% # filter for live trees
  filter(Year %in% c(1994,2019)) %>% #filter for start and end dates
  group_by(TreeID) %>% 
  summarise(N=n()) %>% 
  filter(N==2) %>% #filter for trees with 5 measurements
  pull(TreeID)

@@

# We will produce diameter class stand attributes for two time periods 1992 and 2019.

@@

  # Determine which plots have measurements in 1992 and 2019
  X %>% 
   filter(Year %in% c(1994,2019)) %>% 
   filter(Status=="Live") %>%

@@

By year, mean plot volume, basal area and stand density did not differ
significantly between both harvested units (i.e., p\>0.05,
`r appNums("standFitM.contrasts",display="cite")` and
`r figNums("standPlots",display="cite")`). Contrasts between the
high-removal and control treatment units showed that volume, basal area
and stand density differences were highest and significant (i.e.,
p\<0.05) in the 1994 measurement year, and these differences diminshed
over each subsequent measurement year. For example, mean plot density in
the high-removal treatment unit was approximately

`r standFitM.contrasts %>% filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Stand density") %>%  filter(Year==1994) %>% mutate(estimate=abs(round(estimate,0))) %>%  pull(estimate)`

trees per hectare less than the control in 1994 (p\<0.05); by 2019, that
difference had decreased to

`r standFitM.contrasts %>% filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Stand density") %>%  filter(Year==2019) %>% mutate(estimate=abs(round(estimate,0))) %>%  pull(estimate)`

trees per hectare (p\>0.05). Contrasts between the low-removal and the
control treatment units showed a similar trend with basal area and stand
density; the control unit had significantly higher basal area and stand
density means in 1994, but the difference between the units diminished
over each subsequent measurement, becoming non-significant (i.e.,
p\>0.05) in 2019. Plot means of volume did not differ significantly
between the treatment units in any measurement year.

### Growth dominance statistics

@@

# Filter only for trees that have measurements in 1992 and 2019 
gd.treeList<- # assign variable
  X %>% 
  filter(TreeID!="New") %>% # remove new trees
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(Status=="Live") %>% # filter for live trees
  group_by(TreeID) %>% 
  summarise(N=n()) %>% 
  filter(N>=2) %>% #filter for trees with at least 2 measurements
  pull(TreeID)

@@

Across all plots and time periods, smaller trees contributed
proportionally more to stand basal area growth than larger trees
(`r figNums("gds.plot",display="cite")`). Over the last two measurement
periods, the growth dominance statistic decreased in all plots except
one (not shown), indicating that individual tree growth contributions to
stand growth became more proportional to tree size over the last 11
growing seasons.

@@

A comparison of proportional species composition by diameter class,
treatment and year showed that fir and spruce volume proportion by
diameter class varied similarly between the high RBA and control units
(`r figNums("sppCompDiam.plot",display="cite")`). In 1992, the largest
diameter classes were dominated by spruce, whereas fir was the dominant
species in the smallest diameter classes. This pattern was also found
for basal area and tree density (not shown). By 2019, the fir-spruce
composition by distribution class was maintained in these two units,
with small increases in spruce composition in the smallest diameter
class.

@@

A total of `r deadList %>% nrow()` trees
(`r deadList %>% filter(Species=="Fir") %>% nrow()` fir,
`r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and
`r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died in
`r length(unique(deadList$Plot))` plots from 1994 to 2019. The number of
trees that died per plot from 1994 to 2019 increased significantly with
plot basal area in 1994 (`r figNums("mort.plot",display="cite")`, p \<
0.0001). Mortality was generally not attributable to a particular forest
health factor, and in the majority ofcases, we could not ascertain a
casual agent. Similar linear models fitted separately for fir and spruce
showed that fir mortality had a similar relationship to plot basal area,
whereas spruce mortality did not (not shown). The diameter distribution
of fir mortality followed a negative exponential distribution, with
mortality concentrated in smaller diameter classes
(`r figNums("mortSppDiam.plot",display="cite")`). Spruce mortality was
concentrated in larger diameter classes. By treatment unit, the control
unit had the highest mortality, whereas the high-removal treatment unit
had the lowest mortality (`r figNums("mortDiam.plot",display="cite")`).
The majority of mortality occured during the most recent measurement
period of 2009-2019.

@@

At the tree-level, our results suggest that while both hybrid white
spruce and subalpine fir respond to thinning with faster radial growth,
heavy thinning can elicit a stronger radial growth response in spruce,
compared to fir. Spruce trees in the low-removal harvest treatment did
not respond with a significant growth release. THus, our results suggest
that a relatively high removal of the stand will result in increased
spruce growth, and this release will exceed that of co-occurring fir
trees. Other studies have shown that residual spruce growth responds
favourably to heavy thinning. A 10-year analysis of white spruce growth
underneath an aspen canopy found that a complete or high overstory
removal resulted in immediate and significant spruce radial and height
growth responses, compared to control and partial cutting treatments
(Smith et al. 2016). In a 35-year old white spruce stand, diameter
increment was maximized with the heaviest thinning intensity, however,
the treatment thinning from below (Stiell 1970). Frank (1973) also
showed that 70-year old white spruce basal increment increased with the
amount of thinning. Competition for light is a key factor that limits
white spruce growth, and opening up the stand to allow for increased
light penetration may be a primary mechanism that enhances white spruce
growth responses [@Lieffers1999]. High basal area removal may also
increase air and soil temperatures, which can preferentially favour
spruce growth responses compared to fir [@Lajzerowicz2004]. The
relatively low spruce growth response in the low-removal treatment may
be explained in that the gaps created may have been too small to elicit
as strong a growth response as observed in the high-removal thinning
treatment. Light levels may differ little between forest canopies that
are closed or contain small gaps [@Bartemucci2002], and strong vertical
light gradients may result in relatively little light regime changes for
understorey trees.

In comparison, both partial harvest treatments stimulated a growth
response in fir, compared to the control. Although both species are
similar in shade tolerance, firs are considered to be better adapted to
low light conditions in the understory, allowing them to persist longer
in undisturbed forest (Antos 2000) and potentially able to respond
faster to even modest changes in understory light (Dumais et al 2014,
Messier et al. 1999). Although fir growth may respond at lower levels of
overstorey removal, the growth response potential of fir may be lower
than spruce. Comparative studies of spruce and fir growth rates present
differing evidence regarding species differences in release potential to
thinning. In mature Engelmann spruce (*Picea engelmannii*) -- subalpine
fir forests in western North America, understory spruce growth rates
exceeded those of fir, allowing spruce to achieve canopy position
faster, including after canopy gaps were created through disturbance
(Antos et al. 2000, Andrus et al. 2018). In mature forest in central BC,
spruce growth exceeded that of subalpine fir after age 80 (Eis and
Craigdallie 1983). Because spruce tends to live longer than subalpine
fir, higher growth rates at later ages may allow this species to become
larger than fir. In contrast, thinning studies in spruce-fir forests in
eastern North America report that balsam fir (*Abies balsamifera*)
responds to release with faster height and diameter growth than
co-occurring red spruce (*Picea rubens*) (Dumais and Prevost, 2007;
Dumais et al. 2014, Messier et al. 1999). Other studies have found no
difference in growth response between spruce and fir species after
treatment (MCaughey and Smith1982). Our study also suggests that at
lower levels of basal area removal, spruce and fir growth responses are
similar.

Post-treatment growth rates increased with tree size at the time of
treatment, for both species. This has been found in other studies as
well (references) and may be related to size-asymmetric competition
whereby larger trees preempt resources, especially light, from smaller
individuals [@Bose2017]. Thinning treatments may exacerbate
size-assymetric competition [@Cordonnier2015], and growth increases from
treatment will be predominantly concentrated in the larger tree size
classes [@Bose2018; @Schwinning1998]. Smaller trees will likely continue
to experience competition and will not benefit as much from the thinning
treatment. Thinning treatments may not promote growth responses in the
largest dominant trees in a stand either, as these trees may not
experience significant changes in resource availability or may have
achieved maximal growth rates [@Ryan1997, Bose2017]. Forest managers
should account for potential size differential when planning thinning
treatments. For example, a silviculture treatment that seeks to release
a suppressed understorey may have to ensure that sufficient overstorey
trees are removed to reduce assymetric competition after treatment.

@@

Stand density The thinned treatments increased tree density faster than
the control, and within three decades, there were no significant density
differences between the two harvested treatments and control. This
likely resulted from two processes; growth and mortality. Increasing
stand density reflects the growth of smaller trees that achieved the
minimum diameter class (i.e., 7.5cm DBH) in this study. The faster
increases in stand density of the thinned treatments compared to the
control likely reflect that smaller trees in the thinned stands
responded with faster diameter increment after treatment (as previously
discussed). We note that our study did not account for new seedlings
that may have become established over the period of the study. Based on
increasing densities in the smaller diameter classes, these stands will
likely maintain a negative-exponential diameter distribution, which is
considered typical of older, uneven-aged stands in this region.

@@

Over the 27-year period of this study, plots with high-removal harvest
treatments accrued basal area significantly faster than the control and
low-removal harvest treatment, likely due to increases in tree-level
growth and stand density discussed previously. As a result, differences
in basal area between the high-removal harvest unit and the control
reduced from ca. 17m2/hectare in 1992 to ca.9m2/hectare in 2019.

In contrast, basal area increment between the low-removal unit and the
control did not differ significantly over the 1992-2019 period, and the
difference between these two units remained relatively constant over all
measurement periods at ca. 10m2/hectare. The similar rates of basal area
increment between the low-removal thinning treatment and the control are
difficult to explain. Tree-level results showed that fir grew faster in
the low-removal unit than the control, but spruce did not. As well, by
2019 there were no significant differences in stand density between the
two units. Thus, it appears that while the fir trees may have increased
growth as a response of low-removal, the growth increases were
insufficient to cause a significant change in basal area increment.
Further monitoring in this study will help determine the future
trajectory of the low-removal thinning treatment.

@@

*The dependent variable in the first model (BAI) is tree basal area
increment from 1992-2019. The dependent variable in the second model
(BAI (%)) is basal area increment from 1992-2019 as a percentage of the
tree's basal area in 1992.*

@@

  # plotting
  ggplot()+
  aes(x=Init,y=response,fill=Species,color=Species)+
  geom_point()+
  geom_line()+
  geom_errorbar(aes(ymin=response-SE, ymax=response+SE), width=.2,
                 position=position_dodge(0.05))+
  xlab("1992 DBH (cm)")+
  ylab("Tree BAI from 1992-2019 (m²)")+
  facet_wrap(~BA.Target)+
  theme(legend.position="bottom")

@@

  # plotting
  ggplot()+
  aes(x=Init,y=response,fill=Species,color=Species)+
  geom_point()+
  geom_line()+
  geom_errorbar(aes(ymin=response-SE, ymax=response+SE), width=.2,
                 position=position_dodge(0.05))+
  xlab("1992 DBH (cm)")+
  ylab("Tree BAI from 1992-2019 (% increase)")+
  facet_wrap(~BA.Target)+
  theme(legend.position="bottom")

@@

`r figNums("treeGrowthPct.plot",caption="Estimated marginal means of spruce and fir basal area increment (expressed as a percentage increase relative to 1992 DBH) across five inital DBH classes and three treatments.  Whiskers are standard error of the mean.  Responses are back-transformed from the model. BAI, basal area increment")`

@@

  ggplot()+
  aes(x=Year,y=mean,fill=BA.Target,color=BA.Target,group=BA.Target)+
  geom_point()+
  geom_line()+
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, width=.2))+
                   # position=position_dodge(.9))+
   scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
                     minor_breaks = c(1992,1994,1997,2009,2019),
                     name="Measurement year")+ # stretch out x-axis
  facet_wrap(~Var,scales="free_y")+
  xlab("Plot mean")+
  theme(legend.position = "bottom")

@@

  ggplot()+
  aes(x=Year,y=Mean,group=Species,fill=Species,color=Species)+
  geom_point()+
  geom_line()+
  geom_errorbar(aes(ymin=Mean-SE, ymax=Mean+SE, width=.2),
                   position=position_dodge(0))+
 scale_x_continuous(limits=c(1992,2019),breaks=c(1992,1994,1997,2009,2019),
                     minor_breaks = c(1992,1994,1997,2009,2019),
                     name="Measurement year")+ # stretch out x-axis
  ylab("")+
  xlab("Measurement year")+
  facet_grid(Var~BA.Target,
             scales="free_y",
              switch="y")+
  theme(legend.position = "bottom",
  axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
  strip.placement = "outside",
            strip.background.y = element_blank())

@@

`r appNums("trtContrast.table",caption="Contrasts of predicted 1992-2019 tree growth among treatments within each species.  Ratio of predicted growth between treatments is in 'ratio' column.")`

@@

    # Insert number of growing seasons column
  add_column(Num.Seasons=c(rep(3,6),rep(11,6)),.after="Period") %>% 
  mutate(Period=paste0(Period," (",Num.Seasons,")")) %>% 
  dplyr::select(-Num.Seasons) %>%

    # format table
  flextable() %>%
  align(align="left",part="all") %>% 
  merge_v("Period") %>% 
  valign(valign="top",part="body") %>% 
  hline(i=seq(from=3,to=12,by=3),border=fp_border(color="black")) %>% 
  fix_border_issues() %>% 
   autofit()

>>>>

---
title: "Manuscript_draft"
author: "Michael Jull and Hardy Griesbauer"
date: "10/12/2020"
output:
  word_document:
    reference_docx: wordTemplate2.docx
  html_document:
    df_print: paged
csl: cjfr.csl
bibliography: library.bib
editor_options: 
  markdown: 
    wrap: 72
---

@@

To test the hypothesis that tree-level radial growth response over the
1994-2019 period varied with interactions between (i) tree diameter at
the time of final treatment (1994); (ii) treatment; and (iii) species,
we fit linear mixed effects models with log-transformed 1994-2019
diameter increment as the response variable, and treatment, species and
log-transformed 1994 diameter as interacting fixed effects. To account
for plot-level productivity differences, we included plot as a random
intercept effect. Models were fit using the 'lme4' package for R [\@].
We also fit a similar model with the 1994-2019 diameter increment
expressed as a percentage of 1994 diameter. Plots of residuals and fit
lines with orginal data were examined to evaluate model goodness of fit.
Variance explained for each model are reported using the approach
described in Nakagawa et al. [-@Nakagawa2013], as implemented in the
MuMIn package for R [@barton2009mumin]. Pairwise contrasts were
performed using the 'emmeans' package for R [\@] and Tukey HSD method of
*p* value adjustment was used to control for multiplicity. We report all
significant results at p\<0.05.

@@

# Filter only for trees that have measurements in 1994 and 2019 
treeList<- # assign variable
  X %>% 
  filter(TreeID!="New") %>% # remove new trees
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(Species %in% c("Spruce","Fir")) %>% # filter birch and fdi out of dataset
  filter(Status=="Live") %>% # filter for live trees
  filter(Year %in% c(1994,2019)) %>% #filter for start and end dates
  group_by(TreeID) %>% 
  summarise(N=n()) %>% 
  filter(N==2) %>% #filter for trees with 5 measurements
  pull(TreeID)

@@

# We will produce diameter class stand attributes for two time periods 1994 and 2019.

@@

  # Determine which plots have measurements in 1994 and 2019
  X %>% 
   filter(Year %in% c(1994,2019)) %>% 
   filter(Status=="Live") %>%

@@

By year, mean plot volume, basal area and stand density did not differ
significantly between both harvested units (i.e., p\>0.05,
`r appNums("standFitM.contrasts",display="cite")` and
`r figNums("standPlots",display="cite")`). Contrasts between the
high-removal and control treatment units showed that volume, basal area
and stand density differences were highest and significant (i.e.,
p\<0.05) in the 1994 measurement year, and these differences diminshed
over each subsequent measurement year. For example, mean plot density in
the high-removal treatment unit was approximately
`r standFitM.contrasts %>% filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Stand density") %>%  filter(Year==1994) %>% mutate(estimate=abs(round(estimate,0))) %>%  pull(estimate)`
trees per hectare less than the control in 1994 (p\<0.05); by 2019, that
difference had decreased to 
`r standFitM.contrasts %>% filter(Contrast=="TU:10 - TU:Control") %>% filter(Variable=="Stand density") %>%  filter(Year==2019) %>% mutate(estimate=abs(round(estimate,0))) %>%  pull(estimate)`
trees per hectare (p\>0.05). Contrasts between the low-removal and the
control treatment units showed a similar trend with basal area and stand
density; the control unit had significantly higher basal area and stand
density means in 1994, but the difference between the units diminished
over each subsequent measurement, becoming non-significant (i.e.,
p\>0.05) in 2019. Plot means of volume did not differ significantly
between the treatment units in any measurement year.

### Growth dominance coefficients

@@

# Filter only for trees that have measurements in 1994 and 2019 
gd.treeList<- # assign variable
  X %>% 
  filter(TreeID!="New") %>% # remove new trees
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(Status=="Live") %>% # filter for live trees
  group_by(TreeID) %>% 
  summarise(N=n()) %>% 
  filter(N>=2) %>% #filter for trees with at least 2 measurements
  pull(TreeID)

@@

The growth dominance coefficient was negative for all plots and time periods, indicating that smaller trees contributed
proportionally more to stand basal area growth than larger trees
(`r figNums("gds.plot",display="cite")`). Over the last two measurement
periods, the growth dominance coefficient increased in all plots except
one (not shown), indicating that individual tree growth contributions to
stand growth became more proportional to tree size over the last 11
growing seasons.

@@

A comparison of proportional species composition by diameter class,
treatment and year showed that fir and spruce volume proportion by
diameter class varied similarly between the high RBA and control units
(`r figNums("sppCompDiam.plot",display="cite")`). In 1994, the largest
diameter classes were dominated by spruce, whereas fir was the dominant
species in the smallest diameter classes. This pattern was also found
for basal area and tree density (not shown). By 2019, the fir-spruce
composition by distribution class was maintained in these two units,
with small increases in spruce composition in the smallest diameter
class.

@@

A total of `r deadList %>% nrow()` trees
(`r deadList %>% filter(Species=="Fir") %>% nrow()` fir and
`r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce) died in
`r length(unique(deadList$Plot))` plots from 1994 to 2019. The number of
trees that died per plot from 1994 to 2019 increased significantly with
plot basal area in 1994 (`r figNums("mort.plot",display="cite")`, p \<
0.0001). In the majority of cases, we could not identify a casual agent, and overall, mortality was generally not attributable to a particular forest
health factor. Linear models fitted separately for fir and spruce
showed that fir mortality had a similar relationship to plot basal area,
whereas spruce mortality did not (not shown). The diameter distribution
of fir mortality followed a negative exponential distribution, with
mortality concentrated in smaller diameter classes
(`r figNums("mortSppDiam.plot",display="cite")`). Spruce mortality was
concentrated in larger diameter classes. By treatment unit, the control
unit had the highest mortality, whereas the high-removal treatment unit
had the lowest mortality (`r figNums("mortDiam.plot",display="cite")`).
The majority of mortality occured during the most recent measurement
period of 2009-2019.

@@

At the tree-level, our results suggest that while both hybrid white
spruce and subalpine fir respond to thinning with faster radial growth,
heavy thinning can elicit a stronger radial growth response in spruce,
compared to fir. Spruce trees in the low-removal harvest treatment did
not respond with a significant growth release. Thus, our results suggest
that a relatively high removal of the stand will result in increased
spruce growth, and this release will exceed that of co-occurring fir
trees. Other studies have shown that residual spruce growth responds
favourably to heavy thinning. A 10-year analysis of white spruce growth
underneath an aspen canopy found that a complete or high overstory
removal resulted in immediate and significant spruce radial and height
growth responses, compared to control and partial cutting treatments
(Smith et al. 2016). In a 35-year old white spruce stand, diameter
increment was maximized with the heaviest thinning intensity, however,
the treatment thinning from below (Stiell 1970). Frank (1973) also
showed that basal area increment in a 70-year old white spruce stand increased with the
amount of thinning. Competition for light is a key factor that limits
white spruce growth, and opening up the stand to allow for increased
light penetration may be a primary mechanism that enhances white spruce
growth responses [@Lieffers1999]. High basal area removal may also
increase air and soil temperatures, which can preferentially favour
spruce growth responses compared to fir [@Lajzerowicz2004]. The
relatively low spruce growth response in the low-removal treatment may
be explained in that the gaps created may have been too small to elicit
as strong a growth response as observed in the high-removal thinning
treatment. Light levels may differ little between forest canopies that
are closed or contain small gaps [@Bartemucci2002], and strong vertical
light gradients may result in relatively little light regime changes for
understorey trees.

In comparison, both partial harvest treatments stimulated a growth
response in fir, compared to the control. Firs are considered to be better adapted than spruce to
low light conditions in the understory, allowing them to persist longer
in undisturbed forest (Antos 2000) and potentially able to respond
faster to even modest changes in understory light (Dumais et al 2014,
Messier et al. 1999). Although fir growth may respond at lower levels of
overstorey removal, the growth response potential of fir may be lower
than spruce. Comparative studies of spruce and fir growth rates present
differing evidence regarding genus-level differences in release potential to
thinning. In mature Engelmann spruce (*Picea engelmannii*) - subalpine
fir forests in western North America, understory spruce growth rates
exceeded those of fir, allowing spruce to achieve canopy position
faster, including after canopy gaps were created through disturbance
(Antos et al. 2000, Andrus et al. 2018). In a mature forest in central BC,
spruce growth exceeded that of subalpine fir after age 80 (Eis and
Craigdallie 1983). Because spruce tends to live longer than subalpine
fir, higher growth rates at later ages may allow this species to become
larger than fir. In contrast, thinning studies in spruce-fir forests in
eastern North America report that balsam fir (*Abies balsamifera*)
responds to release with faster height and diameter growth than
co-occurring red spruce (*Picea rubens*) (Dumais and Prevost, 2007;
Dumais et al. 2014, Messier et al. 1999). Other studies have found no
difference in growth response between spruce and fir species after
treatment (MCaughey and Smith1982). Our study also suggests that at
lower levels of basal area removal, spruce and fir growth responses are
similar.

Post-treatment growth rates increased with tree size at the time of
treatment, for both species. This has been found in other studies as
well (references) and may be related to size-asymmetric competition
whereby larger trees preempt resources, especially light, from smaller
individuals [@Bose2017]. Thinning treatments may exacerbate
size-assymetric competition [@Cordonnier2015], and growth increases from
treatment will be predominantly concentrated in the larger tree size
classes [@Bose2018; @Schwinning1998]. Smaller trees will likely continue
to experience competition and will not benefit as much from the thinning
treatment. Thinning treatments may not promote growth responses in the
largest dominant trees in a stand either, as these trees may not
experience significant changes in resource availability or may have
achieved maximal growth rates [@Ryan1997, Bose2017].

Although growth responses were concentrated in the larger trees, our analysis of growth dominance indicated that smaller trees contributed proportionally more to stand growth than larger trees.  Binkley et al. (2006) propose that GDC in developing forests increases from 0 to positive during early development, reflecting the growth dominance of larger robust trees as canopy closure is achieved.  As the forests age, the GDC may switch to negative, reflecting several co-occuring processes.  Older trees may start to senesce, with growth reductions and increased mortality.  Younger trees may increase their resource use efficiency, and/or increase their allocation to leaf biomass (Baret et al. 2017) partially in response to increasing resource availiabilty as older trees senesce.  Data from old Englemann spruce – subalpine fir forests in Colorado (Binkley et al. 2006) as well as old-growth boreal mixedwood stands (Baret et al. 2017) support this proposed model of growth dominance development.  While we found a similar negative growth dominance in the stands of this study, a comparison between the two recent measurement periods showed that growth dominance actually increased over time (i.e., tree growth became increasingly proportional to tree size).  This finding is unique from other studies, and suggests that either larger trees are increasing their growth rates relative to smaller trees, or smaller trees are beginning to experience increased competition and may be experiencing lower growth rates.  Further data from these stands will help evaluate the long-term trend in GDC.

Our results suggest that thinning did not substantially alter the growth dominance of the stand within close to 30 years after treatment.  Similar results were reported after thinning in uneven- and even-aged hardwood stands in eastern Canada [@Lemire2020].  While thinning treatments stimulated a growth response in all trees, the increase in growth was more or less proportional to tree size.  A significant decrease in GDS could be potentially achieved in older stands where the treatment seeks to release suppressed larger trees [@Lemire2020].  Conversely, thinning even-aged red pine (*Pinus resinosa*)  stands reduced growth dominance compared to unthinned stands, regardless of thinning type, presumably by allocating more growing space to previously suppressed trees [@Bradford2010].  The use of GDC to evaluate the effects of partial cutting may be insufficient on its own, and should be used in conjunction with tree-level analyses [@Lemire2020].  Forest managers should account for potential size differential when planning thinning
treatments. For example, a silviculture treatment that seeks to release
a suppressed understorey may have to ensure that sufficient overstorey
trees are removed to reduce assymetric competition and growth dominance after treatment.

@@

The thinned treatments increased tree density faster than
the control, and within three decades, there were no significant density
differences between the two harvested treatments and control. This
likely resulted from two processes; growth and mortality. Increasing
stand density reflects the growth of smaller trees that achieved the
minimum diameter class (i.e., 7.5cm DBH) in this study. The faster
increases in stand density of the thinned treatments compared to the
control likely reflect that smaller trees in the thinned stands
responded with faster diameter increment after treatment (as previously
discussed). We note that our study did not account for new seedlings
that may have become established over the period of the study. Based on
increasing densities in the smaller diameter classes, these stands will
likely maintain a negative-exponential diameter distribution, which is
considered typical of older, uneven-aged stands in this region.

@@

Over the 27-year period of this study, plots with high-removal harvest
treatments accrued basal area significantly faster than the control and
low-removal harvest treatment, likely due to increases in tree-level
growth and stand density discussed previously. As a result, differences
in basal area between the high-removal harvest unit and the control
reduced from ca. 17m2/hectare in 1994 to ca.9m2/hectare in 2019.

In contrast, basal area increment between the low-removal unit and the
control did not differ significantly over the 1994-2019 period, and the
difference between these two units remained relatively constant over all
measurement periods at ca. 10m2/hectare. The similar rates of basal area
increment between the low-removal thinning treatment and the control are
difficult to explain. Tree-level results showed that fir grew faster in
the low-removal unit than the control, but spruce did not. As well, by
2019 there were no significant differences in stand density between the
two units. Thus, it appears that while the fir trees may have increased
growth as a response of low-removal, the growth increases were
insufficient to cause a significant change in basal area increment.
Further monitoring in this study will help determine the future
trajectory of the low-removal thinning treatment.

@@

*The dependent variable in the first model (BAI) is tree basal area
increment from 1994-2019. The dependent variable in the second model
(BAI (%)) is basal area increment from 1994-2019 as a percentage of the
tree's basal area in 1994.*

@@

  # plotting
  ggplot()+
  aes(x=Init,y=response,fill=Species,color=Species)+
  geom_point()+
  geom_line()+
  geom_errorbar(aes(ymin=response-SE, ymax=response+SE), width=.2,
                 position=position_dodge(0.05))+
  xlab("1994 DBH (cm)")+
  ylab("Tree BAI from 1994-2019 (m²)")+
  facet_wrap(~BA.Target)+
  theme(legend.position="bottom")

@@

  # plotting
  ggplot()+
  aes(x=Init,y=response,fill=Species,color=Species)+
  geom_point()+
  geom_line()+
  geom_errorbar(aes(ymin=response-SE, ymax=response+SE), width=.2,
                 position=position_dodge(0.05))+
  xlab("1994 DBH (cm)")+
  ylab("Tree BAI from 1994-2019 (% increase)")+
  facet_wrap(~BA.Target)+
  theme(legend.position="bottom")

@@

`r figNums("treeGrowthPct.plot",caption="Estimated marginal means of spruce and fir basal area increment (expressed as a percentage increase relative to 1994 DBH) across five inital DBH classes and three treatments.  Whiskers are standard error of the mean.  Responses are back-transformed from the model. BAI, basal area increment")`

@@

  ggplot()+
  aes(x=Year,y=mean,fill=BA.Target,color=BA.Target,group=BA.Target)+
  geom_point()+
  geom_line()+
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, width=.2))+
                   # position=position_dodge(.9))+
   scale_x_continuous(limits=c(1994,2019),breaks=c(1994,1997,2009,2019),
                     minor_breaks = c(1994,1997,2009,2019),
                     name="Measurement year")+ # stretch out x-axis
  facet_wrap(~Var,scales="free_y")+
  xlab("Plot mean")+
  theme(legend.position = "bottom")

@@

  ggplot()+
  aes(x=Year,y=Mean,group=Species,fill=Species,color=Species)+
  geom_point()+
  geom_line()+
  geom_errorbar(aes(ymin=Mean-SE, ymax=Mean+SE, width=.2),
                   position=position_dodge(0))+
 scale_x_continuous(limits=c(1994,2019),breaks=c(1994,1997,2009,2019),
                     minor_breaks = c(1994,1997,2009,2019),
                     name="Measurement year")+ # stretch out x-axis
  ylab("")+
  xlab("Measurement year")+
  facet_grid(Var~BA.Target,
             scales="free_y",
              switch="y")+
  theme(legend.position = "bottom",
  axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
  strip.placement = "outside",
            strip.background.y = element_blank())

@@

`r appNums("trtContrast.table",caption="Contrasts of predicted 1994-2019 tree growth among treatments within each species.  Ratio of predicted growth between treatments is in 'ratio' column.")`

@@

    # Insert number of growing seasons column
  add_column(Num.Seasons=c(rep(3,3),rep(11,6)),.after="Period") %>% 
  mutate(Period=paste0(Period," (",Num.Seasons,")")) %>% 
  dplyr::select(-Num.Seasons) %>%

    # format table
  flextable() %>%
  align(align="left",part="all") %>% 
  merge_v("Period") %>% 
  valign(valign="top",part="body") %>% 
  hline(i=seq(from=3,to=9,by=3),border=fp_border(color="black")) %>% 
  fix_border_issues() %>% 
   autofit()
